Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-08-30 writeup/conStruct.tex
complete draft

\begin{abstract}
One of the classic problems in population genetics is the characterization 
of discrete population structure when the genotyped samples also show 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
clustering or assignment methods may incorrectly ascribe differentiation 
due to continuous processes (e.g., isolation by distance) 
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This is partly a result of the difficulty of sampling uniformly and continuously 
across the range of a population or species, 
but more, it reflects a shortcoming of current methods for inferring and 
visualizing population structure from genetic data in the face of data 
that are characterized by both continuous and discrete population structure.
Here, we present a novel statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of discrete population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This model explicitly addresses the ``clines vs. clusters" problem in 
modeling population genetic variation by jointly accommodating both 
continuous and discrete patterns of differentiation. 
We demonstrate the utility of this approach using both simulations and an empirical application.
\end{abstract}

@@

%In studying populations we need group individuals 
%Many downstream

%Biology diversity

%The fundamental goal of evolutionary genetics is to link patterns of genetic variation 
%to the processes that have generated and maintained them.
%Often, the first step in the analysis of genetic data is to define evolutionary units of analysis
%by delineating discrete populations from which individuals or alleles are sampled \citep{wright1949genetical}.
%These units are useful for defining a scope of study in research on selection or local adaptation (ref),
%or for identifying either locally adapted alleles (human skin color, ralph \& coop, gartersnakes) or alleles involved in disease (cystic fibrosis).
%Delineating populations can also be useful for informing conservation priorities \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.
%However, if underlying population structure is continuous, this may entail a loss of power.

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \citep{CavalliSforza1975, treemix},
$k$-means clustering analyses performed on dimensionality-reduction approaches (ref), 
such as principal components analysis \citep{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches \citep[e.g.][]{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, FINESTRUCTURE, fastStructure, huelsenbeck2007inference, Corander2003,TESS,geneland}, among others.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Like the method we present here, these are based on a statistical model,
but do not perform inference directly with a demographic model.
Each of these methods perform best in particular situations,
but all can give misleading results when applied to data that show a continuous pattern of isolation by distance 
\citep{Wright1943, novembre_interpreting_2008, Frantz2009}.

@@

To address this, we set out to develop
a model-based clustering method that when possible uses isolation by distance 
to explain observed genetic variation.
With an explicit spatial component, discrete population structure need only be invoked when genetic differentiation 
in the data is deviates significantly from that expected given geographic separation.

@@

In addition to these differences on the diagonal,
this differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eqn. \eqref{allelic_covariance} further in the Appendix (\secref{allelic_cov}).
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $D_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Sigma_{i,j} = (1 - 2 D_{i,j})/4$.
Therefore, unlike genetic covariance,
this allelic covariance is affected in shape by singleton sites,
so it may be advisable to filter these prior to analysis
if these are likely to contain a large percentage of errors.

\paragraph{Continuous and discrete differentiation}
Clustering approaches to describing genetic variation
are useful because population history can often be meaningfully described on a coarse scale
by interactions between discrete ``populations''
whose relationships are delimited by patterns of glaciation,
large-scale migration,
mountain ranges, and the like.
Here we add a spatial component within each such discrete historical component,
which we refer to as a set of ``layers'' which overlay the modern map.
We imagine each layer as a geographically distributed population,
and that each sample is composed of a mixture of contributions from each of these layers.
These layers thus take the place of ``clusters'' or ``populations'' in clustering methods,
but we do not adopt these terms as they connote a lack of geographical separation within each.

Within each of these layers,
allele frequencies have positive covariance at geographically close locations,
but this covariance is allowed to decay towards zero as geographic distance increases.
This pattern of spatial decay reflects how migration between neighboring demes  
homogenizes allele frequency changes that arise locally due to drift, 
but this mixing less effectively homogenizes geographic distant demes,
resulting in a continuous pattern of isolation by distance within each cluster.
There is a fixed amount of covariance between layers, irrespective of spatial location.
Within each layer allele frequencies are expected to change gradually with distance,
but observed frequencies can change abruptly at many loci 
if the proportions by which individuals' ancestry derive from each layer 
(the ``admixture'' coefficients) 
do so as well.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_cluster_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is strongly conceptually similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomial distributed (see XXXX).  
Therefore, we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework.

\paragraph{Choice of cluster number and cross-validation}
There are a number of reasons, both statistical and biological, 
why there is no true (or right) number of layers for real datasets,
discussed further in the Discussion.
However, it is still important to assess whether additional spatial layers (larger $K$)
meaningfully model pattern in the data
or merely explain spurious variation introduced by noise
-- in other words, whether additional model complexity
provides significant explanatory power.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} results across different values of $K$ 
and between the spatial and nonspatial models.

Several approaches have been used as model choice criteria 
for the number of clusters in population genetic data, including: 
the use of information theoretic criteria \citep[e.g.][]{ADMIXTURE}, such as AIC or BIC;
comparisons of the likelihood of the data across different values of $K$, 
with various criteria on how to choose a single value (e.g., \citep{Evanno2005});
or comparisons of the marginal likelihood, 
generated either via various approximations \citep[e.g.][]{STRUCTURE}
or via a more rigorous and computationally intensive approach such as thermodynamic integration \citep{verity_nichols2016}
or inference using a Dirichlet process prior \citep{huelsenbeck2007inference}
See \cite{verity_nichols2016} for a discussion of these approaches and comparison
between several methods.

We choose to take a cross-validation approach \citep[similar in spirit to][]{ADMIXTURE_xval}.
We estimate the posterior on the parameters of the model using a ``training" partition of the data 
(in practice the training set is a random subset of $90\%$ of the loci). 
We then calculate the likelihood of the ``testing" subset (the remaining $10\%$ of the loci), 
averaged over the posterior of the parameters. 
We repeat this procedure a number of times, 
take an average of the relative log-likelihoods over the runs at each $K$, 
and select the model with the significantly best predictive accuracy
(see \secref{Xvalidation} for more on our cross-validation procedure).
This approach rewards models with larger values of $K$, 
which increase the likelihood of the training partition, 
as they fit the data better,
but the improvement in the likelihood will plateau (or peak), 
as above a certain $K$ the model is fitting noise specific to the training data 
and therefore not fitting the testing set any better.
\plr{I wonder if we can just cite \cite{verity_nichols2016} rather than this laundry list here?}

We use cross-validation to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

\gb{something about how we choose the best model?}
\plr{how's this?}
Cross-validation provides a valuable summary of how much explanatory power
is added by spatial structure within each layer, and each additional layer.
A simple $t$-test can be used to assess whether additional layers
show statistically significantly greater predictive accuracy.
However, we remind users that ``statistical significance does not imply real-world significance'',
and so small but statistically significant differences between models
should probably not be relied on too strongly.

@@

We ran \texttt{conStruct} analyses on each of these simulated datasets, 
specifying both a spatial and nonspatial model for $K = $ 1 through 7, 
then used cross-validation to compare the predictive performance of the models.
In Fig \ref{simK1_pies}, we show the results of the nonspatial model 
run with between 1 and 7 clusters, applied to the dataset simulated under $K=1$,
and in Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
The admixture pie plots for the spatial and nonspatial models run on the datasets simulated under $K=2$ and $K=3$
are shown in Figs \ref{simK1_nsp_pies}-\ref{simK3_sp_pies}.
To confirm that results from our nonspatial model were qualitatively similar to those of 
other nonspatial model-based clustering algorithms, 
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, 
inferring admixture under models with $K=2$ through 4.

A number of results are immediately clear from these tests with simulated data.
First, \texttt{conStruct} performs well on all simulated scenarios.
Across all values of K and for all simulations, 
the spatial model is correctly preferred over the nonspatial model, 
and it infers the true admixture proportions with high accuracy, tight precision, and good coverage 
(Fig \ref{sim_xvals}).
Second, the predictive accuracy of the nonspatial model increases 
with the number of clusters in the model, 
as it describes continuous, spatial patterns of relatedness between demes within the same cluster 
using gradients of admixture across more and more discrete clusters.
This behavior of the nonspatial model is highlighted in 
the analyses of data simulated under $K=1$ (Fig \ref{simK1_pies} and \ref{simK1_nsp_pies}):
as $K$ in the model increases, 
the sampled area is increasingly divvied up between clusters 
with ``centers of mass" in the $K$ geographical extremes of the simulated space.
At $K=2$, the clusters are centered in opposite corners on the map; 
at $K=3$, in the lower two corners and the top half; 
at $K=4$, all four corners, and so on.
These results are qualitatively similar to those generated using fastSTRUCTURE \citep{fastStructure}, 
which also describes continuous differentiation using admixture between clusters
 (Fig \ref{fastStr_simK1}-\ref{fastStr_simK3}).
While it is true that each such cluster is more similar within itself
than it is to other clusters,
we know that these boundaries are arbitrary,
since the data were simulated with no such boundaries.

In contrast, the predictive accuracy of the spatial \texttt{conStruct} model
either plateaus at or declines from the true value of $K$ used to simulate the data (Fig \ref{sim_xvals}, second column).
And, as $K$ in the spatial model increases, 
deme membership in layers beyond the number used to simulate the data does not appreciably increase 
(Fig \ref{simK1_sp_pies}, \ref{simK2_sp_pies}, \ref{simK3_sp_pies}).
That is, even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.

\gb{could talk about proportion of relatedness explained by each layer here.
e.g., Another way to visualize the difference in the behavior of the spatial and nonspatial models 
is to summarize the proportion of relatedness between samples explained by each cluster...}

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{simK1_nsp_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_nsp_pies_K2.pdf}}
		\subcaptionbox{$K=3$\label{simK1_nsp_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_nsp_pies_K3.pdf}}
		\subcaptionbox{$K=4$\label{simK1_nsp_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_nsp_pies_K4.pdf}}
		\subcaptionbox{$K=2$\label{simK1_sp_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_sp_pies_K5.pdf}}
		\subcaptionbox{$K=3$\label{simK1_sp_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_sp_pies_K6.pdf}}
		\subcaptionbox{$K=4$\label{simK1_sp_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_sp_pies_K7.pdf}}
	\caption{
	Map of admixture proportions estimated using the nonspatial model for $K=2$ through 4 (top row)
	and also the spatial model for $K=2$ through 4 (bottom row).
	The data were simulated using a single cluster with nearest-neighbor symmetric migration between demes.
    }\label{simK1_pies}
\end{figure}

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=8.5in]{figs/sims/xvals.pdf}}
	\caption{
	Cross-validation results for data simulated under $K=1$, $K=2$, and $K=3$, 
	comparing the spatial and nonspatial \texttt{conStruct} models run with $K=1$ through 7.  
	The first panel in each row shows all results; 
	the second panel shows results for just the spatial model.
	The inset plots zoom in on cross-validation results outlined in the dotted boxes.
    }\label{sim_xvals}
\end{figure}

@@

We address these questions using a dataset generated and originally published by \cite{geraldes_etal_2014}.
The published data consist of 434 individuals sampled from 35 drainages 
and genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera} individuals).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data, grouped by drainage, using both the spatial and nonspatial \texttt{conStruct} models, 
using $K = 1$ through 7,
and compared these models using cross-validation.
The results of these analyses are shown in 
Figs \ref{populus_xvals}--\ref{populus_sp_clst_covs} and \ref{populus_nsp_clst_covs}. %,populus_nsp_pies,populus_sp_clst_covs,populus_nsp_clst_covs

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=4in]{figs/populus/populus_sp_clst_covs.pdf}}
	\caption{
	Plots showing the cluster-specific parametric covariance IBD-curves 
	estimated for the \textit{Populus} data using 
	the spatial \texttt{conStruct} model run with $K=2$ through 7.
	Line colors are consistent with cluster colors in Fig \ref{populus_sp_pies}.
	Points are colored by the species they are a covariance between:
	black on black points are sample covariances between populations of \textit{Populus balsamifera};
	green on black points are sample covariances between \bals{} and \tri{};
	green on green points are sample covariances between \tri{} and \tri{}.
    }\label{populus_sp_clst_covs}
\end{figure}

@@

A consistent split between layers within \tri{} fell along the ``no-cottonwood belt," 
a region along the central coast of British Columbia in which black cottonwood is absent. 
The no-cottonwood belt is hypothesized to divide the species' distribution 
into northern and southern groups, which, in a provenance test, 
were experimentally shown to display different phenotypes 
\citep[pathogen resistance][]{xie2009,xie2012}.  
At higher values of $K$, drainages at the southern tip of \tri{} sampling 
begin to split out into their own layers.  
\plr{cut this if we pick lower $K$? Or add a disclaimer?}
This may potentially be the result of introgression from Southern \textit{Populus} species, 
such as \textit{P. angustifolia} or \textit{fremontii} \citep{Zhou2012,geraldes_etal_2014}

At all values of $K$, the spatial model was preferred over the nonspatial model.
However, most of this signal seems to derive from the \textit{P. trichocarpa} samples:
covariance within predominantly \tri{} layers decays more quickly with distance
than within the majority \bals{} layer, as seen in Fig \ref{populus_sp_clst_covs}.
The most strongly supported model, given the current dataset, was XXXX.

\subsubsection*{Black bears}

>>>>

\begin{abstract}
One of the classic problems in population genetics is the characterization 
of discrete population structure when the genotyped samples also show 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
clustering or assignment methods may incorrectly ascribe differentiation 
due to continuous processes (e.g., isolation by distance) 
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This is partly a result of the difficulty of sampling uniformly and continuously 
across the range of a population or species, 
but more, it reflects a shortcoming of current methods for inferring and 
visualizing population structure from genetic data in the face of data 
that are characterized by both continuous and discrete population structure.
Here, we present a novel statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of discrete population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This model explicitly addresses the ``clines vs. clusters" problem in 
modeling population genetic variation by jointly accommodating both 
continuous and discrete patterns of differentiation. 
We demonstrate the utility of this approach using both simulations and empirical applications.
\end{abstract}

@@

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \citep{CavalliSforza1975, treemix},
$k$-means clustering analyses performed on dimensionality-reduction approaches \citep{meirmans2009genodive}, 
such as principal components analysis \citep{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches \citep[e.g.][]{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, FINESTRUCTURE, fastStructure, huelsenbeck2007inference, Corander2003,TESS,geneland}, among others.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Like the method we present here, these are based on a statistical model,
but do not perform inference directly with a demographic model.
Each of these methods perform best in particular situations,
but all can give misleading results when applied to data that show a continuous pattern of isolation by distance 
\citep{Wright1943, novembre_interpreting_2008, Frantz2009}.

@@

To address this, we set out to develop
a model-based clustering method that, when possible, uses isolation by distance 
to explain observed genetic variation.
With an explicit spatial component, discrete population structure need only be invoked when genetic differentiation 
in the data deviates significantly from that expected given geographic separation.

@@

In addition to these differences on the diagonal,
this differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eqn. \eqref{allelic_covariance} further in the Appendix (\secref{allelic_cov}).
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $D_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 D_{i,j})/4$.
Therefore, unlike genetic covariance,
this allelic covariance is affected in shape by singleton sites,
so it may be advisable to filter these prior to analysis
if these are likely to contain a large percentage of errors.

\paragraph{Continuous and discrete differentiation}
Clustering approaches to describing genetic variation
are useful because population history can often be meaningfully described on a coarse scale
by interactions between discrete ``populations''
whose relationships are delimited by patterns of glaciation,
large-scale migration,
mountain ranges, and the like.
Here we add a spatial component within each such discrete historical component,
which we refer to as a set of ``layers'' that overlay the modern map.
We imagine each layer as a geographically distributed population,
and that each sample is composed of a mixture of contributions from each of these layers.
These layers thus take the place of ``clusters'' or ``populations'' in clustering methods,
but we do not adopt these terms as they connote a lack of geographical separation within each.

Within each of these layers,
allele frequencies have positive covariance at geographically close locations,
but this covariance is allowed to decay as geographic distance increases.
This pattern of spatial decay reflects how migration between neighboring demes  
homogenizes allele frequency changes that arise locally due to drift, 
but less effectively homogenizes geographic distant demes,
resulting in a continuous pattern of isolation by distance within each cluster.
There is a fixed amount of covariance between layers, irrespective of spatial location.
Within each layer allele frequencies are expected to change gradually with distance,
but observed frequencies can change abruptly at many loci 
if the proportions by which individuals' ancestry derive from each layer 
(the ``admixture'' coefficients) 
do so as well.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_cluster_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is strongly conceptually similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
Therefore, we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework.

\paragraph{Choice of cluster number and cross-validation}
There are a number of reasons, both statistical and biological, 
why there is no true (or right) number of layers for real datasets,
discussed further in the Discussion.
However, it is still important to assess whether additional spatial layers (larger $K$)
meaningfully model patterns in the data
or merely explain spurious variation introduced by noise
-- in other words, whether additional model complexity
provides significant explanatory power.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} results across different values of $K$ 
and between the spatial and nonspatial models.

Several approaches have been used as model choice criteria 
for the number of clusters in population genetic data, including: 
information theoretic criteria \citep[e.g.][]{ADMIXTURE}, such as AIC or BIC;
comparisons of the likelihood of the data across different values of $K$, 
with various criteria on how to choose a single value (e.g., \citep{Evanno2005});
or comparisons of the marginal likelihood, 
generated either via various approximations \citep[e.g.][]{STRUCTURE}
or via a more rigorous and computationally intensive approach such as thermodynamic integration \citep{verity_nichols2016}
or inference using a Dirichlet process prior \citep{huelsenbeck2007inference}.
See \cite{verity_nichols2016} for a discussion of these approaches and comparison
between several methods.

We use cross-validation \citep[similar in spirit to][]{ADMIXTURE_xval} to attack this problem.
To do this,
we use a ``training" partition of the data (in practice, a random 90\% subset of the loci)
to estimate the posterior distribution of the parameters,
and then calculate the log-likelihood of the remaining ``testing'' loci,
averaged over the posterior.
Prediction accuracy of a particular model (e.g., value of $K$)
is then measured using this log-likelihood,
averaged over a number of independent data partitions.
The best model is judged to be the simplest one with significantly better predictive accuracy
than others (see \secref{Xvalidation} for more on our cross-validation procedure).
In general, larger values of $K$ allows the model more flexibility,
and thus increases the likelihood of the training partition, 
but this improvement in the likelihood will plateau (or even peak), 
since above a certain $K$ the model only fits noise specific to the training data 
rather than generalizable patterns.
At any value of $K$, support for the spatial model over the nonspatial model 
means that isolation by distance is likely a feature of the data.

Cross-validation provides a valuable summary of how much explanatory power
is added by spatial structure within each layer, and each additional layer.
A simple $t$-test can be used to assess whether additional layers
show statistically significantly greater predictive accuracy.
However, we remind users that ``statistical significance does not imply real-world significance'',
and so small but statistically significant differences between models
should probably not be relied on too strongly.

An additional heuristic to evaluate the inclusion of additional layers is to calculate 
each layer's relative contribution to total covariance as their importance, 
and to choose a value of $K$ where all layers have an importance above some cutoff (e.g. 0.1\%).
Because the prior on admixture proportions is a Dirichlet distribution 
that is quite harsh against intermediate admixture values (see Table \ref{tab:param_prior_tab}),  
the model will tend to ``not use" unnecessary layers, 
and they will have a low contribution to overall covariance.
The contribution of the $k$th layer can be calculated as
its contribution to the parametric covariance between samples 
summed over the upper triangle (excluding the diagonal) of the covariance matrix.
The importance of the $k$th layer is its contribution relative to those of all other layers, 
i.e., its contribution divided by the sum of the contributions of all $K$ layers
(for details on calculating this importance, see Appendix \secref{layer_importance}). 
This layer importance is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

We ran \texttt{conStruct} analyses on each of these simulated datasets, 
specifying both a spatial and nonspatial model for $K = $ 1 through 7, 
then used cross-validation to compare the predictive performance of the models.
In Fig \ref{simK1_pies}, we show the results of the analyses of the dataset simulated under $K=1$, 
contrasting the results of the nonspatial model runs with 2 - 4 clusters 
with those of spatial model with 2 - 4 layers.
In Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
The results of all analyses run on datasets simulated using $K=1$ through 3 are shown in 
Figs \ref{simK1_nsp_pies}-\ref{simK3_adprop_fit}.
To confirm that results from our nonspatial model were qualitatively similar to those of 
other nonspatial model-based clustering algorithms, 
we also analyzed each simulated dataset using fastSTRUCTURE \citep{fastStructure}, 
inferring admixture under models with $K=2$ through 4 (Figs \ref{fastStr_simK1}-\ref{fastStr_simK3}).

A number of results are immediately clear from these tests with simulated data.
First, \texttt{conStruct} performs well on all simulated scenarios.
Across all values of K and for all simulations, 
the spatial model is correctly preferred over the nonspatial model (Fig \ref{sim_xvals})., 
and it infers the true admixture proportions with high accuracy, tight precision, and good coverage 
(Figs \ref{simK2_adprop_fit} and \ref{simK3_adprop_fit}).
Second, the predictive accuracy of the nonspatial model increases 
with the number of clusters in the model, 
as it describes continuous, spatial patterns of relatedness between demes within the same cluster 
using gradients of admixture across more and more discrete clusters.
This behavior of the nonspatial model is highlighted in 
the analyses of data simulated under $K=1$ (Fig \ref{simK1_pies} and \ref{simK1_nsp_pies}):
as $K$ in the model increases, 
the sampled area is increasingly divvied up between clusters 
with ``centers of mass" in the $K$ geographical extremes of the simulated space.
At $K=2$, the clusters are centered in opposite corners on the map; 
at $K=3$, in the lower two corners and the top half; 
at $K=4$, all four corners, and so on.
These results are qualitatively similar to those generated using fastSTRUCTURE \citep{fastStructure}, 
which also describes continuous differentiation using admixture between clusters
 (Fig \ref{fastStr_simK1}-\ref{fastStr_simK3}).
While it is true that each cluster is more similar within itself
than it is to other clusters,
we know that these boundaries are arbitrary,
since the data were simulated with no such boundaries.

In contrast, the predictive accuracy of the spatial \texttt{conStruct} model
either plateaus at or declines from the true value of $K$ used to simulate the data 
(Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}).
And, as $K$ in the spatial model increases, 
deme membership in layers beyond the number used to simulate the data does not appreciably increase 
(Fig \ref{simK1_sp_pies}, \ref{simK2_sp_pies}, \ref{simK3_sp_pies}).
That is, even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
These results are highlighted in the calculation of layer importance for 
analyses run in each simulation (Fig \ref{simK1_laycon}, \ref{simK2_laycon}, and \ref{simK3_laycon}).
In the spatial analyses run using the true value of $K$,
the importance of the $K$ layers remains constant even with additional layers;
in contrast, additional layers in the nonspatial analyses continue to be important for describing variation.

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{simK1_pies:nsp_K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_nsp_pies_K2.pdf}}
		\subcaptionbox{$K=3$\label{simK1_pies:nsp_K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_nsp_pies_K3.pdf}}
		\subcaptionbox{$K=4$\label{simK1_pies:nsp_K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_nsp_pies_K4.pdf}}
		\subcaptionbox{$K=2$\label{simK1_pies:sp_K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_sp_pies_K5.pdf}}
		\subcaptionbox{$K=3$\label{simK1_pies:sp_K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_sp_pies_K6.pdf}}
		\subcaptionbox{$K=4$\label{simK1_pies:sp_K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/sims/simK1_sp_pies_K7.pdf}}
	\caption{
	Map of admixture proportions estimated using the nonspatial model for $K=2$ through 4 (top row)
	and also the spatial model for $K=2$ through 4 (bottom row).
	The data were simulated using a single cluster with nearest-neighbor symmetric migration between demes.
    }\label{simK1_pies}
\end{figure}

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=2in]{figs/sims/sim_xvals.pdf}}
	\caption{
	Cross-validation results for data simulated under $K=1$, $K=2$, and $K=3$, 
	comparing the spatial and nonspatial \texttt{conStruct} models run with $K=1$ through 7.  
	The inset plots zoom in on cross-validation results outlined in the dotted boxes.
    }\label{sim_xvals}
\end{figure}

@@

Our data, originally published by \cite{geraldes_etal_2014}, 
consist of 434 individuals sampled from 35 drainages 
and genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera} individuals).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data, grouped by drainage, using both the spatial and nonspatial \texttt{conStruct} models, 
using $K = 1$ through 7,
and compared these models using cross-validation.
The results of these analyses are shown in 
Figs \ref{populus_xvals} and \ref{populus_pies}, 
as well as in Figs \ref{populus_sp_pies} - \ref{populus_laycon} in the Supplementary Materials.

@@

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{populus_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp2.pdf}}
		\subcaptionbox{$K=3$\label{populus_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp3.pdf}}
		\subcaptionbox{$K=4$\label{populus_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp4.pdf}}
	\caption{
	Map of admixture proportions estimated for the \textit{Populus} dataset 
	using the spatial model for $K=2$ through 4.
    }\label{populus_pies}
\end{figure}

@@

A consistent split between layers within \tri{} fell along the ``no-cottonwood belt," 
a region along the central coast of British Columbia in which black cottonwood is absent. 
The no-cottonwood belt is hypothesized to divide the species' distribution 
into northern and southern groups, which, in a provenance test, 
were experimentally shown to display different phenotypes 
\citep[e.g., pathogen resistance,][]{xie2009,xie2012}.  
At higher values of $K$, drainages at the southern tip of \tri{} sampling 
begin to split out into their own layers.  
This may potentially be the result of introgression from southern \textit{Populus} species, 
such as \textit{P. angustifolia} or \textit{fremontii} \citep{Zhou2012,geraldes_etal_2014}

At all values of $K$, the spatial model was preferred over the nonspatial model (Fig \ref{populus_xvals}).
Most of this signal seems to derive from the \textit{P. trichocarpa} samples:
covariance within predominantly \tri{} layers decays more quickly with distance
than within the majority \bals{} layer, as seen in Fig \ref{populus_sp_clst_covs}.  
This indicates that, although isolation by distance is a feature of the dataset as a whole, 
it's magnitude varies considerably across biological units - 
an aspect of the data \texttt{conStruct} is able capture well.

\subsubsection*{Black bears}
The American black bear, \textit{Ursus americanus}, is endemic to North America
and has a broad distribution across the continent.
During the last glacial maximum, 
black bears were confined to isolated glacial refugia, 
from which they subsequently expanded to occupy their current range
\citep{WoodingWard1997,Byun1997,Stone2000,Puckett2015}.
The retreat of the glaciers was relatively recent in units of 
black bear generation times \citep{Onorato2004}, 
so we might reasonably expect the divergence accrued between refugia 
to be reflected in discrete genetic structure between extant populations.
Black bears are strong dispersers \citep{Moore2014}, 
but they are distributed across most of the continent, 
so we may therefore expect them to experience distance-limited 
mating opportunities and exhibit a pattern of IBD.
As with the poplars described above, 
this system is likely to present patterns of genetic structure 
that are both continuous and discrete, 
making it well-suited as a test for our method.
We organize the results of our analyses of 
the complicated landscape of genetic structure in black bears 
around the following questions:
\begin{enumerate}
\item How many historical clusters are reflected in modern patterns of genetic variation?
\item Do we see support for a pattern of isolation by distance within each of the inferred discrete groups?
\end{enumerate}

Our data, originally published by \cite{Puckett2015}, consist of 95 individuals 
sampled across the United States and on the West coast of Canada,
genotyped at just under 22,000 bi-allelic loci (map of the sampling shown in Fig \ref{bear_map}).
The distribution of missing data across these individuals was uneven, 
with a few individuals representing most of the missing data, 
so we removed individuals with greater than 4\% missing data, 
resulting in a final dataset of 78 individuals.
We then analyzed these data, treating individuals as the unit of analysis, 
using both the spatial and nonspatial \texttt{conStruct} models with a $K$ of between 1 and 7, 
and compared these models using cross-validation.
The results of these analyses are shown in Figs \ref{bear_xvals} - \ref{bear_laycon}, 
as well as in Figs \ref{bear_sp_pies} - \ref{bear_nsp_clst_covs} in the Supplementary Materials.

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=4in]{figs/bears/bear_sampling_map.pdf}}
	\caption{
	Map of the sampled black bear individuals included in the analysis.
    }\label{bear_map}
\end{figure}
